Tropical Mathematics and the Lambda Calculus

### 2023-01-23 appIV.tex
Commit notturno (finita parte Taylor)

@@

\begin{proposition}
The category $\LREL_!$, endowed with the operator $D$, is a cartesian closed differential category.
\end{proposition}
\begin{proof}
We must check axioms (D1)-(D7) of cartesian differential categories  plus axiom (D-curry) (cf.~\cite{Manzo2012}).
\begin{description}
\item[(D1)]$D(\min\{f,g\})=\min \{D(f),D(g)\}$ and $D(\infty)=\infty$: 
while the latter is obvious, for the former we have 
$D(\min\{f,g\})_{[x]\oplus\nu,y}= \min\{f,g\}_{\nu+x,y}= \min\{f_{\nu+x,y},g_{\nu+x,y}\}=\min\{D(f),D(g)\}_{[x]\oplus \nu,y}$, and if $\mu\neq [x]\oplus \nu$, 
$D(\min\{f,g\})_{\mu,y}= \infty= \min\{\infty, \infty\}=\min\{ D(f),D(g)\}_{\mu,y}$. 
\item[(D2)]
$D(f)\circ \langle \min\{h,k\},v\rangle= \min\{ D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\}$, and $D(f)\circ \langle \infty,v\rangle=\infty$: we can compute
	\begin{align*}
	(D(f)\circ \langle \min\{h,k\},v\rangle)_{\mu,y}&=
	\inf\Big\{ 
	\sum_{i=1}^{n}\min\{h,k\}_{\rho_{i},w_{i}}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\\
&	\qquad\qquad\mid
	\mu=\sum_{i=1}^{n}\rho_{i}+\sum_{j=1}^{m}\nu_{j},	[w]=[w_{1},\dots, w_{n}]
	\Big\}\\
	& 
=	\inf\Big\{ 
	\min\{h,k\}_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\\
&	\qquad\qquad\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
	\Big\}\\
	&=	\min\Big\{\inf\big\{ 
	h_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}, \\
	& \qquad\qquad \inf\big\{ 
	k_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}\Big\}\\
	& = \min\Big\{(D(f)\circ \langle h,v\rangle)_{\mu,y}, (D(f)\circ \langle k,v\rangle)_{\mu,y}\Big\}\\
		& = \Big(\min\big\{D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\big\}\Big )_{\mu,y}
	\end{align*}
	where, in the first equation, the condition $[w_{1},\dots, w_{n}]=[w]$ (i.e.~$n=1$) is forced by the fact that, otherwise, the application of $D(f)$ would give $\infty$. Moreover, we have
\begin{align*}
(D(f) \circ \langle \infty, v\rangle)_{\mu,y}&=\inf\Big\{ 
	\infty+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
	\Big\} = \infty
\end{align*}

@@

\paragraph{Taylor Expansion }

To check the validity of Taylor expansion we must further check the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
$$
\mathsf{ev}\circ \langle f,g\rangle= \inf_{n\in \BB N}
(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{n\text{ times}})\circ \langle \mathrm{id}, \infty\rangle
$$
where:
\begin{enumerate}
\item $\mathsf{ev}\in \LREL_!(B^{A}+A, B)$ is the canonical \emph{evaluation} morphism;

@@

\subsubsection{The language and the semantics of $\STDLC$}

The language we mainly consider in the main paper il the differential $\lambda$-calculus $\STDLC$, that we quickly define here, following [Section 3, \cite{Manzo2010}].

The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars, then quotiented by usual $\alpha$-equivalence plus other equivalences handling the $+$ constructor.

{
\begin{minipage}{\textwidth}
\begin{align*}
\textit{Terms: } & M,N:= x\mid \lambda x.M \mid M\BB T \mid \Der [M] \cdot N
\\
\textit{Sums: } & \BB T:= 0 \mid M \mid M+\BB T
\end{align*}\end{minipage}}\medskip\\
Simple types are generated by the grammar:

@@

We finally recall the sound interpretation of $\STDLC$-terms in a $C\partial C$ category, following [Section 4.3, \cite{Manzo2010}].

\begin{definition}
Let $\C C$ be a $C\partial C$ and $\model X$ fixed objects of $\C C$, one for each ground type $X$ of the $\STDLC$.

@@

One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{ER}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).

@@

>>>>

\subsubsection{Proof of Theorem~\ref{thm:LREL!CCDC}}

Theorem~\ref{thm:LREL!CCDC} says that $(\LREL_!,D)$ is a $CC\partial C$.
This is proved in Proposition~\ref{propApp:CCDC}.
First we need the following:

@@

\begin{proposition}\label{propApp:CCDC}
The category $\LREL_!$, endowed with the operator $D$, is a cartesian closed differential category.
\end{proposition}
\begin{proof}
We must check axioms (D1)-(D7) of cartesian differential categories  plus axiom (D-curry) (cf.~\cite{Manzo2012}).
\begin{description}
\item[(D1)]$D(\min\{f,g\})=\min \{D(f),D(g)\}$ and $D(\infty)=\infty$: 
while the latter is obvious, for the former we have 
$D(\min\{f,g\})_{[x]\oplus\nu,y}= \min\{f,g\}_{\nu+x,y}= \min\{f_{\nu+x,y},g_{\nu+x,y}\}=\min\{D(f),D(g)\}_{[x]\oplus \nu,y}$, and if $\mu\neq [x]\oplus \nu$, 
$D(\min\{f,g\})_{\mu,y}= \infty= \min\{\infty, \infty\}=\min\{ D(f),D(g)\}_{\mu,y}$. 
\item[(D2)]
$D(f)\circ \langle \min\{h,k\},v\rangle= \min\{ D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\}$, and $D(f)\circ \langle \infty,v\rangle=\infty$: we can compute
	\begin{align*}
	(D(f)\circ \langle \min\{h,k\},v\rangle)_{\mu,y}&=
	\inf\Big\{ 
	\sum_{i=1}^{n}\min\{h,k\}_{\rho_{i},w_{i}}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\\
&	\qquad\qquad\mid
	\mu=\sum_{i=1}^{n}\rho_{i}+\sum_{j=1}^{m}\nu_{j},	[w]=[w_{1},\dots, w_{n}]
	\Big\}\\
	& 
=	\inf\Big\{ 
	\min\{h,k\}_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\\
&	\qquad\qquad\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
	\Big\}\\
	&=	\min\Big\{\inf\big\{ 
	h_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}, \\
	& \qquad\qquad \inf\big\{ 
	k_{\rho,w}+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}	\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}\big\}\Big\}\\
	& = \min\Big\{(D(f)\circ \langle h,v\rangle)_{\mu,y}, (D(f)\circ \langle k,v\rangle)_{\mu,y}\Big\}\\
		& = \Big(\min\big\{D(f)\circ \langle h,v\rangle, D(f)\circ \langle k,v\rangle\big\}\Big )_{\mu,y}
	\end{align*}
	where, in the first equation, the condition $[w_{1},\dots, w_{n}]=[w]$ (i.e.~$n=1$) is forced by the fact that, otherwise, the application of $D(f)$ would give $\infty$. Moreover, we have
\begin{align*}
(D(f) \circ \langle \infty, v\rangle)_{\mu,y}&=\inf\Big\{ 
	\infty+
	\sum_{j=1}^{m}v_{\nu_{j},z_{j}}
	+
	f_{[z_{1},\dots, z_{m}]+w,y}\mid
	\mu=\rho+\sum_{j=1}^{m}\nu_{j}
	\Big\} = \infty
\end{align*}

@@

%\paragraph{Taylor Expansion }
\subsubsection{Taylor Expansion in $\LREL_!$}

To check the validity of Taylor expansion in $\LREL_!$ we must further check the following equation, given $f\in \LREL_!(C, B^{A})$ and $g\in \LREL_!(C,A)$:
$$
\mathsf{ev}\circ \langle f,g\rangle= \inf_{n\in \BB N}
\left\{(( \cdots (\Lambda^{-}(f) \underbrace{\star g)\cdots )\star g}_{n\text{ times}})\circ \langle \mathrm{id}, \infty\rangle\right\}
$$
where:
\begin{enumerate}
\item $\mathsf{ev}\in \LREL_!(B^{A}+A, B)$ is the canonical \emph{evaluation} morphism;

@@

\subsubsection{The language and the semantics of the $\STDLC$}\label{secApp:STDLC}

The language we mainly consider in the main paper is the differential $\lambda$-calculus $\STDLC$, that we briefly define here, following [Section 3, \cite{Manzo2010}].

The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars:

{
\begin{minipage}{\textwidth}
\begin{align*}
\textit{Terms: } & M,N:= x\mid \lambda x.M \mid M\BB T \mid \Der [M] \cdot N \qquad \textit{(Call $\Lambda^\partial$ the set of such terms)}
\\
\textit{Sums: } & \BB T:= 0 \mid M \mid M+\BB T
\end{align*}\end{minipage}}\medskip\\
quotiented by usual $\alpha$-equivalence plus other equivalences.
Among those, we ask that $+$ is commutative, associative, with neutral element $0$ (thus the name ``sums''), and the constructors $\lam x.(\cdot)$, $(\cdot)\BB T$ and $\Der [\cdot] \cdot (\cdot)$ commute with $+$.
One says that those constructors are \emph{linear}.
Remark that $M(\cdot)$ is \emph{not} set to be linear.
So, while $\lambda x.0=0\BB T=\Der [0] \cdot N=\Der [M] \cdot 0=0$, we have $M0\neq0$ in general.
This is crucial for the definition of the Taylor expansion.
One also needs to quotient for an equation, called ``permutative equality'', which basically says that the order of \emph{consecutive} linear applications $\Der [\cdot] \cdot (\cdot)$ does not count.
Finally, for simplicity we follow here the tradition of ``eliminating the coefficients'', that is, we quotient for the idempotency of the sum.
Sums are, then, just (\emph{finite}) subsets of $\Lambda^\partial$.

Simple types are as usual generated by the grammar:

@@

We finally recall the sound interpretation of $\STDLC$-terms in a $CC\partial C$ category, following [Section 4.3, \cite{Manzo2010}].

\begin{definition}
Let $\C C$ be a $CC\partial C$ and $\model X$ fixed objects of $\C C$, one for each ground type $X$ of the $\STDLC$.

@@

One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{difflambda}), and thanks to the fact that $\C C$ is a $CC\partial C$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).

@@

\subsubsection{The Taylor expansion of the $\STLC$}

In the last lines of Section~\ref{sec:3B} we state Corollary~\ref{cor:T(M)=M} which is about the Taylor expansion of a ordinary $\lambda$-terms $M$.
In the main paper we do not define it, let us briefly recall it here.
For details, the reader can look at the many papers on the subject pointed in the bibliography.

If sums (as defined in the above lines of Section~\ref{secApp:STDLC}) are \emph{finite} subsets of $\Lambda^\partial$, the Taylor expansion considers in general infinite subsets of $\Lambda^\partial$.
Since we decided to ``eliminate the coefficients'', what we will define is often called the ``qualitative'' Taylor expansion.

\begin{definition}

 The (qualitative) \emph{Taylor expansion} is the following inductively defined map $\Theta : \Lambda \to \C P\left(\Lambda^\partial\right)$.
\[\begin{array}{rcl}
 \Te x & := & x \\
 \Te{\lambda x.M} & := & \lambda x.\Te M \\
 \Te{M\BB T} & := & \sum\limits_{n\in\N} \left(\Der^n [\Te M] \cdot \Te{\BB T}^n\right)0
\end{array}\]
where $\Te{M+\BB T}:=\Te M+\Te{\BB T}$ and $\Der^{n+1} [\Te M] \cdot \Te N^{n+1}:=\Der[\Der^n [\Te M] \cdot \Te N^n] \cdot \Te N$.
\end{definition}

We can now prove Corollary~\ref{cor:T(M)=M}.

\begin{theorem}[Corollary~\ref{cor:T(M)=M}]
 In $(\LREL_!,D)$, we have:
 \[ \model M = \model{\Te M} \]
for all $M\in\Lambda$.
\end{theorem}
\begin{proof}
 By induction on $M$.
 We only give the crucial case, that is $M=P\left(\sum\limits_j N_j\right)$.
Looking at the definition of $\model{\cdot}$ we have:
 \[\begin{array}{rclr}
	\model{P(\sum\limits_j N_j)} 
	& = &
	 \mathsf{ev} \circ \langle \, \model{P}, \model{\sum\limits_j N_j}  \, \rangle
	&
	\\
	& = &
	 \inf\limits_{k\in\N}\left\{
	 ((\dots (\Lambda^-(\model{P})\star \model{\sum\limits_j N_j})\cdots ) \star \model{\sum\limits_j N_j} )
	 \circ \langle id,\infty \rangle \right\}
	&
	\textit{(Theorem~\ref{thm:modelsTaylor})}
	\\
	& = &
	 \inf\limits_{k\in\N}\left\{
	 ((\dots (\Lambda^-(\model{\Te P})\star \model{\sum\limits_j \Te N_j})\cdots ) \star \model{\sum\limits_j \Te N_j} )
	 \circ \langle id,\infty \rangle \right\}
	&
	\textit{(ind.\ hyp.)}
	\\
	& = &
	 \inf\limits_{k\in\N}\left\{
	 \mathsf{ev} \circ \langle\, \Lambda\left(
	 (\dots (\Lambda^-(\model{\Te P})\star \model{\sum\limits_j \Te N_j})\cdots ) \star \model{\sum\limits_j \Te N_j}
	 \right) , \infty \, \rangle \right\}
	& \textit{(general property of CCC's)}
	\\
	& = &
	 \inf\limits_{k\in\N}\left\{
	 \mathsf{ev} \circ \langle\, \model{\Der^n [\Te P] \cdot \Te{\sum\limits_j N_j}^n}, \infty \, \rangle \right\}
	&
	\\
	& = &
	 \inf\limits_{k\in\N}\left\{\model{(\Der^n [\Te P] \cdot \Te{\sum\limits_j N_j}^n)0} \right\}
	&
	\\ \\
	& = &
	 \model{\Te{P(\sum\limits_j N_j)}}
	&
 \end{array}\]
\end{proof}

### 2023-01-23 section2.tex
Commit notturno (finita parte Taylor)

Another intriguing similarity between program derivatives and 
actual derivatives is provided by the \emph{Taylor expansion}:
in $\STDLC$ one can prove that any application $MN$ can be expanded as an infinite formal sum of \emph{linear} applications
$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
\begin{align}\label{eq:taylor}
MN  =  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
\end{align}
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

>>>>

Another intriguing similarity between program derivatives and 
actual derivatives is provided by the \emph{Taylor expansion} $\C T$:
in $\STDLC$ one can expand any application $MN$ as an infinite formal sum of \emph{linear} applications
$\Der^{(k)}[M]\cdot N^{k}$, i.e.~where $N$ is passed exactly $k$ times to $M$:
\begin{align}\label{eq:taylor}
\Te{MN} :=  \sum_{k=0}^{\infty}\frac{1}{!k}\cdot (\Der^{(k)}[M] \cdot N^{k})0
\end{align}
In other words, unbounded duplications correspond to some sort of limit of bounded, but arbitrarily large, ones.% (this perspective is made clearer by the related approach of the \emph{resource $\lambda$-calculus} \cite{}).

### 2023-01-23 section3C.tex
Commit notturno (finita parte Taylor)

\begin{theorem}\label{thm:modelsTaylor}
 In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have that, for all
$t\in\HOM{\LREL_!}{Z}{X\multimap Y}$ and $s\in\HOM{\LREL_!}{Z}{X}$:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align}
 \end{theorem}
%It is worth discussing the formula above a bit more. 
Here,
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to linear application of $s$ to $u$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
version of the usual Taylor expansion.
Indeed, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
%
% corresponding to the syntactical \emph{differential substitution} of $\STDLC$.
%%\begin{proof}
% {\color{red}(Che ci scriviamo ??}
%\end{proof}

>>>>

\begin{theorem}\label{thm:modelsTaylor}
 In $\LREL_!$ equipped with $D$, all morphisms can be Taylor expanded, i.e.\ in $\LREL_!$ we have that, for all
$t\in\HOM{\LREL_!}{Z}{X\multimap Y}$ and $s\in\HOM{\LREL_!}{Z}{X}$:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align}
 \end{theorem}
%It is worth discussing the formula above a bit more. 
Here,
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
Now, since $\LREL_!$ has countable sums (i.e.~all $\inf$'s converge), an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] shows:

\begin{corollary}\label{cor:T(M)=M}
 The interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
\end{corollary}
